\large Neural, Tree, and Symbolic Optimization of Cosmological Particle-Mesh Simulation \vspace{0.3em}

### 2023-07-20 docs/papers/sto/so.tex
FIXUP draft minor
FIXUP draft: add some notes
FIXUP draft: minor edits

\begin{abstract}

@@

  snapshots. Performance difference between scan and while?

@@

\item (paper) Think about what figures and tables to show in the paper. Update

@@

\vspace{1em}
\section{Introduction}

@@

\begin{table*}
\centering
\caption{Ranges of \GADGET\ and \pmwd\ configuration and cosmological
parameters.
Note that the grid ratio need next fast len to determine the mesh shape
for fast FFT.
Given the box size, the mesh shape determines the cell size.
The softening parameter gives the ratio of the comoving softening length
to the mean particle spacing.
The curvature $\Omega_k$ is related to the separate universe simulation.
We sample parameters applicable to \GADGET\ during data generation, and
those applicable to \pmwd\ during training.
\textsuperscript\dag{}The box size is determined jointly by the \pmwd\
mesh shape and mesh cell size below, which are assumed to be sampled
independently. \YL{I forgot why I wrote ``Mpc'' without the $h$ here.}
In practice, we sample one conditioned on the other and the box size.
*Number of time steps from $a=1/16$ to $a=1$.
}
\label{tab:param}
\begin{tabular}{lcccr}
\toprule
applicability & parameter & distribution & range & sampling \\
\midrule
& $128^3$ white noise & $\mathcal{N}(\vzero, \vI)$ & 512 realizations & MC \\
\cmidrule(lr){2-5}
& 121 snapshots, $a$ & uniform & [1/16, 1] & deterministic \\
\cmidrule(lr){2-5}
& box size in Mpc\textsuperscript\dag & log-trapezoidal & [25.6, 2560) \\
\cmidrule(lr){2-4}
& snapshot offset, $\Delta\!a$ & uniform & [0, 1/128) \\
\cmidrule(lr){2-4}
\GADGET\ & $\As \times 10^9$ & log-uniform & [1, 4) \\
\cmidrule(lr){2-4}
\& \pmwd\ & $\ns$ & log-uniform & [0.75, 1.25) & RQMC \\
\cmidrule(lr){2-4}
& $\Omegam$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
\cmidrule(lr){2-4}
& $\Omegab / \Omegam$ & log-uniform & [1/8, 1/4) \\
\cmidrule(lr){2-4}
& $\Omega_k / (1 - \Omega_k)$ & uniform & [-1/3, 1/3) \\
\cmidrule(lr){2-4}
& $h$ & log-uniform & [0.5, 1) \\
\cmidrule(lr){1-4}
\GADGET\ & softening ratio & log-uniform & [1/50, 1/20) \\
\cmidrule(lr){1-5}
& mesh shape & log-uniform & [128, 512] \\
\cmidrule(lr){2-4}
\pmwd\ & mesh cell size in Mpc & log-uniform & [0.2, 5] & MC \\
\cmidrule(lr){2-4}
& number of time steps* & log-uniform & [10, 1000] \\
\cmidrule(lr){1-4}
training & loss grid offset in its own spacing & uniform & $[0, 1)^3$ \\
\bottomrule
\end{tabular}
\end{table*}

\vspace{1em}
\subsection{Parameters and Configurations}

We run 512 \GADGET\ simulations that cover a wide range of
configurations and cosmological parameters, as listed in
\autoref{tab:param}.
Each simulation has $128^3$ particles and writes 121 snapshots linearly
spaced in scale factor from $1/16 + \Delta a$ to $1 + \Delta a$, with
spacing $1/128$.
$\Delta a$ is an offset, common for all snapshots in each simulation,
that varies as in \autoref{tab:param} and helps to sample different
alignment between snapshot and start/stop times.

@@

\vspace{1em}
\subsection{Target Data for Training and Testing}

\citet{GADGET-4}

@@

Given 512 \GADGET\ simulations and 121 output snapshots each, in total
we have 61952 snapshots for training.
Following \autoref{tab:param}, in each iteration, we randomly sample the
snapshots, the number of time steps, and the mesh shape (or its cell
size), and configure \pmwd\ with the same initial coniditions,
cosmological parameters and other common configurations.
To balance the work of GPUs during training, in each iteration we use
the same snapshot $a$, number of time steps, and mesh shape on all GPU.

We perform parallel training using distributed GPU devices.
The training on each snapshot is performed on a single GPU device.
During each training step, the gradients are collected and averaged across
devices to update the SO parameters.

Hyperparameters:

>>>>

\begin{abstract}
% up to 150 words

@@

  snapshots. Performance difference between scan and while?\newline
  \YZ{It seems that lax.while\_loop only supports fwd grad}
  \href{https://jax.readthedocs.io/en/latest/notebooks/Common_Gotchas_in_JAX.html#summary}{link}

@@

\item (paper) Think about what figures and tables to show in the paper. Up

@@

% Main text: up to 3500 words: introduction + results + discussion
% so we should keep introduction and discussion brief
\vspace{1em}
\section{Introduction}
% without heading

@@

\YZ{We further sum the loss of all snapshots across a single forward simulation
at different times, i.e.
\begin{equation}
\cJ = \sum_{a_i} \cJ |_{a_i}
\end{equation}
}

\begin{table*}
\centering
\caption{Ranges of \GADGET\ and \pmwd\ configuration and cosmological
parameters.
Note that the grid ratio need next fast len to determine the mesh shape
for fast FFT.
Given the box size, the mesh shape determines the cell size.
The softening parameter gives the ratio of the comoving softening length
to the mean particle spacing.
The curvature $\Omega_k$ is related to the separate universe simulation.
We sample parameters applicable to \GADGET\ during data generation, and
those applicable to \pmwd\ during training.
\textsuperscript\dag{}The box size is determined jointly by the \pmwd\
mesh shape and mesh cell size below, which are assumed to be sampled
independently. \YL{I forgot why I wrote ``Mpc'' without the $h$ here.}
In practice, we sample one conditioned on the other and the box size.
*Number of time steps from $a=1/16$ to $a=1+1/128$ (in order to cover the
snapshot offset).
}
\label{tab:param}
\begin{tabular}{lcccr}
\toprule
applicability & parameter & distribution & range & sampling \\
\midrule
& $128^3$ white noise & $\mathcal{N}(\vzero, \vI)$ & 512 realizations & MC \\
\cmidrule(lr){2-5}
& 121 snapshots, $a$ & uniform & [1/16, 1] & deterministic \\
\cmidrule(lr){2-5}
& box size in Mpc\textsuperscript\dag & log-trapezoidal & [25.6, 2560) \\
\cmidrule(lr){2-4}
& snapshot offset, $\Delta\!a$ & uniform & [0, 1/128) \\
\cmidrule(lr){2-4}
\GADGET\ & $\As \times 10^9$ & log-uniform & [1, 4) \\
\cmidrule(lr){2-4}
\& \pmwd\ & $\ns$ & log-uniform & [0.75, 1.25) & RQMC \\
\cmidrule(lr){2-4}
& $\Omegam$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
\cmidrule(lr){2-4}
& $\Omegab / \Omegam$ & log-uniform & [1/8, 1/4) \\
\cmidrule(lr){2-4}
& $\Omega_k / (1 - \Omega_k)$ & uniform & [-1/3, 1/3) \\
\cmidrule(lr){2-4}
& $h$ & log-uniform & [0.5, 1) \\
\cmidrule(lr){1-4}
\GADGET\ & softening ratio & log-uniform & [1/50, 1/20) \\
\cmidrule(lr){1-5}
& mesh shape & log-uniform & [128, 512] \\
\cmidrule(lr){2-4}
\pmwd\ & mesh cell size in Mpc & log-uniform & [0.2, 5] & MC \\
\cmidrule(lr){2-4}
& number of time steps* & log-uniform & [10, 1000] \\
\cmidrule(lr){1-4}
training & loss grid offset in its own spacing & uniform & $[0, 1)^3$ \\
\bottomrule
\end{tabular}
\end{table*}

\vspace{1em}
\subsection{Target Data and Configurations}

We generate our target data for training and testing with the \GADGET\
simulation code \citep{GADGET-4}.
We run 512 \GADGET\ simulations that cover a wide range of
configurations and cosmological parameters, as listed in
\autoref{tab:param}.
Each simulation has $128^3$ particles and writes 121 snapshots linearly
spaced in scale factor from $1/16 + \Delta a$ to $1 + \Delta a$, with
spacing $1/128$.
$\Delta a$ is an offset, common for all snapshots in each simulation,
that varies as in \autoref{tab:param} and helps to sample different
alignment between snapshot and start/stop times.

@@

@@

Following \autoref{tab:param}, in each iteration, we randomly sample the number
of time steps and the mesh shape (or its cell size), and configure \pmwd\ with
the same initial coniditions, cosmological parameters and other common
configurations.
To balance the work of GPUs during training, in each iteration we use
the same number of time steps and mesh shape on all GPUs.

We perform parallel training using distributed GPU devices.
The training on all snapshots of a single Sobol configuration is performed on a
single GPU device.
In each training step, the loss and gradients are collected and averaged across
devices.
The SO parameters are initialized to be the same for all devices, and being
updated with the same averaged gradients.

Hyperparameters:
\YZ{
\begin{itemize}
  \item Model: number of hidden layers and nodes of f and g nets
  \item Optimization: learning rate, weight decay and dropout rate etc
\end{itemize}
}
